---
title: "DS 202 Final Project"
author: "Kellan Bouwman, Adam Ford, Heather Moore, Brett White, Ashley Sackpraseuth, Felix Baez"
output: html_document
---
## Voter Participation Analysis by Congression District for the 2018 Midterm Election

**Overview**
    This dataset comes from the Census Bureau, it is collected data for the 2018 midterm election. They have the overall population of people who are of Voting age within the district as well as the number of votes cast. In other tables they have data on that districts population such as age, education, sex, poverty status, race, and geographic area. We are looking at those specifics to see what has the strongest corrolation for voter paricpation.

# Introduction & Cleaning
# *Brett White*
    For this section....
     
```{r}
# tidyverse
library(tidyverse)
# readxl
library(readxl)
df1 <- read.csv('table01.csv')
df2 <- read.csv('table02a.csv')
df3 <- read.csv('table02b.csv')
df4 <- read.csv('table02c.csv')
df5 <- read.csv('table02d.csv')
census_race <- read.csv("census_race_data.csv")
# head(df1)
# head(df2)
# head(df3)
# head(df4)
# head(df5)
#head(census_race_data)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(lubridate)
library(sf)
#install.packages(gridExtra)
library(gridExtra)
library(ggthemes)
```

# Education Analysis
# *Kellan Bouwman*
    When looking at the specific educational categories there does not seem to be any specific trend. If we are lumping them into a group though, the more educated they are the higher voting percentage for a congressional district. This is further seen when looking at Highschool and bachelor degree estimates compared with the national average, the more educate a district is the more participation there is. With Highschool graduates the estimate range has a lot more variability, but has a similar estimating ability as bachelor degrees. 

```{r}
eduPol <- df4 %>% select(State.abbreviation, State.name, Congressional.district, Citizen.voting.age.population.estimate, Less.grade.9.estimate, Grade.9.12.no.diploma.estimate, High.school.graduate.estimate, Some.college.no.degree.estimate, Associate.degree.estimate, Bachelor.degree.estimate, Graduate.degree.estimate, High.school.or.more.estimate, Bachelor.degree.or.more.estimate)

eduPol %>% full_join(df1) -> eduPol

eduPol %>% pivot_longer(Less.grade.9.estimate:Graduate.degree.estimate, names_to = "Education", values_to = "Amount") -> eduPol
#head(eduPol)

eduPol %>%
  mutate(PercentageEdu = (Amount / Citizen.voting.age.population.estimate) * 100 ) -> eduPol
```
    This gives us the national average for highest level of education. This is nice to see for when we are looking at comparsion graphs to see how the national average compares to the correlation graph.
```{r}
eduPol %>%
  group_by(Education) %>%
  summarise(mean = mean(PercentageEdu, na.rm=TRUE))
```
    While the table above gives us the national averages for highest level of education, we should also look more broad categories such on average in a district how many people have a high school diploma or more education as well as who has at least a bachelors degree or more.
```{r}
highSchoolPlus <- mean((eduPol$High.school.or.more.estimate / eduPol$Citizen.voting.age.population.estimate) * 100)
BachPlus <- mean((eduPol$Bachelor.degree.or.more.estimate / eduPol$Citizen.voting.age.population.estimate) * 100)
VoteringRateAvg <- mean(eduPol$Voting.rate.estimate, na.rm = TRUE)
print(cat("The national average for those who have at least highschool level education or more is: ", highSchoolPlus, "%, and the national average per congresional district the percent of people who have at least a bachelors degree is: ", BachPlus, "%, finally the national average voting rate per district is ", VoteringRateAvg, "%"))
```
    Below are graphs of voter participation by precentage of a district with a certain education level. This show us that the higher percentage of your population with a high level of education the voter participation increases. So as the lower highest level of education decreases, the voter participation increases.
```{r}
eduPol %>%
  mutate(nationalAvgDiffHighGrad = ((High.school.or.more.estimate / Citizen.voting.age.population.estimate) * 100) - highSchoolPlus, nationalAvgDiffBach = ((Bachelor.degree.or.more.estimate / Citizen.voting.age.population.estimate) * 100) - BachPlus, highschoolgradper = ((High.school.or.more.estimate / Citizen.voting.age.population.estimate) * 100), bachgradper = ((Bachelor.degree.or.more.estimate / Citizen.voting.age.population.estimate) * 100)) -> eduPol

#Overall Comparisons
ggplot(eduPol, aes(x=PercentageEdu, y=Voting.rate.estimate)) + geom_point() + geom_smooth(method='lm') + facet_wrap(~Education, scales = "free") + ggtitle("Voter Rate based off Highest level of education") + xlab("Percentage of Population") + ylab("Voter Participation Percentage")
```
    With the knowledge that as higher percentage of a district is educated I decided to look at the more general categories to see the trends. So I compared the a district to the national average for a district for at least high school level education obtained and the effect on voter participation. As expect as the rate is above the national average there is an increase in voter participation, and when is it lower there is a voter voting participation. There is a clustering of data near the 0 - 5 percent difference, for at least high school education. There are two possible reasons for this the primary being a ceiling effect at 100%, since the national average is 90.004%, the second is that since the value is so large there are other factors playing a role in voter participation. The affect of these factors led me to look at people with at least a bachelors degrees. 
```{r}
#Highschool Graduation corralation with voter rate
ggplot(eduPol, aes(x=nationalAvgDiffHighGrad, y=Voting.rate.estimate)) + geom_point() + geom_smooth(method='lm') + ggtitle("Voter Rate & difference Highschool or more education & the National Average") + xlab("Percentage of Population, difference with the National Average") + ylab("Voter Participation Percentage")
```
    Then I used a similar graph as the difference of national average of high school with the bachelor degree difference with the national average for bachelor degree's per district. The national average for this is 30.37% which could present a floor effect at 0, but thankfully we do not have any such outliers. Instead even with some percentages being lower than the national average, we start to see a more spread out and constant and better predictor of voter participation. This also further proves the assumption earlier that the higher percent of a district population with higher levels of education. 
```{r}
#Bachelors Degrees holders
ggplot(eduPol, aes(x=nationalAvgDiffBach, y=Voting.rate.estimate)) + geom_point() + geom_smooth(method='lm') + ggtitle("Voter Rate & difference Bachelors Degrees or more education & National Average") + xlab("Percentage of Population, difference with the National Average") + ylab("Voter Participation Percentage")
```

# Age Analysis
# *Adam Ford*

```{r}
table01 <- data.frame(df1)
table01 <- table01 %>%
  rename(
    stateAbb = State.abbreviation,
    stateName = State.name,
    congressDist = Congressional.district,
    votesCastForCongress = Votes.cast.for.congressional.representative.for.the.November.6..2018.election,
    votingAgePopulation = Citizen.voting.age.population.estimate,
    votingAgePopulationMOE = Citizen.voting.age.population.Margin.of.Error,
    votingRate = Voting.rate.estimate,
    votingRateMOE = Voting.rate.Margin.of.Error
  )
head(table01, 5)
table02a <- read.csv("table02a.csv")
table02a <- table02a %>%
  rename(
    stateAbb = State.abbreviation,
    stateName = State.name,
    congressDist = Congressional.district,
    votingAgePopulation = Citizen.voting.age.population.estimate,
    votingAgePopulationMOE = Citizen.voting.age.population.MOE
  )
table02b <- data.frame(df3)
table02b <- table02b %>%
  rename(
    stateAbb = State.abbreviation,
    stateName = State.name,
    congressDist = Congressional.district,
    votingAgePopulation = Citizen.voting.age.population.estimate,
    votingAgePopulationMOE = Citizen.voting.age.population.MOE
  )

table02a <- table01 %>%
  select(stateAbb, congressDist, votingRate) %>%
  inner_join(table02a, by = c('stateAbb', 'congressDist'))
mean18.29 <- mean(table02a$Ages.18.29.percent.estimate)
mean30.44 <- mean(table02a$Ages.30.44.percent.estimate)
mean45.64 <- mean(table02a$Ages.45.64.percent.estimate)
mean65.up <- mean(table02a$Ages.65.up.percent.estimate)
table02a$dif18.29 <- mean18.29 - table02a$Ages.18.29.percent.estimate
table02a$dif30.44 <- mean30.44 - table02a$Ages.30.44.percent.estimate
table02a$dif45.64 <- mean45.64 - table02a$Ages.45.64.percent.estimate
table02a$dif65.up <- mean65.up - table02a$Ages.65.up.percent.estimate
```

### 18 - 29
```{r}
table02a %>% 
  select(stateName, congressDist, dif18.29, votingRate) %>%
  arrange(desc(dif18.29)) %>%
  head(5)
```

### 30 - 44
```{r}
table02a %>%
  select(stateName, congressDist, dif30.44, votingRate) %>%
  arrange(desc(dif30.44)) %>%
  head(5)
```

### 45 - 64
```{r}
table02a %>%
  select(stateName, congressDist, dif45.64, votingRate) %>%
  arrange(desc(dif45.64)) %>%
  head(5)
```

### 65 and up
```{r}
table02a %>%
  select(stateName, congressDist, dif65.up, votingRate) %>%
  arrange(desc(dif65.up)) %>%
  head(5)
```

  These tables show the congressional districts that have the nationwide greatest percentage of voters in the stated age groups. A notable state on these lists is Florida, it dominates the maximums for young voters. A pattern is beginning to show in the Voting Rates as well. We see that the states with high populations of young voters also have a High Voting Rate, and the inverse for high populations of older voters.
  
```{r, message=FALSE, warning=FALSE}
ggplot(table02a, aes(x=dif18.29, y=votingRate)) +geom_point() + geom_smooth(method='lm') + xlab("Difference in Percent Composition of Voters Ages 18-29 compared to the National Average") + ylab("Voting Rate")+ ggtitle("Difference in the 18-29 age group versus Voting Rate in Congressional Districts")

ggplot(table02a, aes(x=dif45.64, y=votingRate)) +geom_point() + geom_smooth(method='lm')  + xlab("Difference in Percent Composition of Voters Ages 45-64 compared to the National Average") + ylab("Voting Rate")+ ggtitle("Difference in the 45-64 age group versus Voting Rate in Congressional Districts")
```

  These two graphs show two starkly different relationships between Difference of Composition of Voters Ages 18-29 and 45-64 versus the respective Voting Rates. In more layman's terms, congressional districts that has a higher percentage of voters 18-29 than the national average, tend to have better voter turnouts. The opposite can be said about the age group of 45-64. Therefore, we can conclude that citizens 18-29 are the group that has the greatest influence on increasing the voting rate.
  
```{r}
table02aByState <- table02a %>%
  group_by(stateName) %>%
  summarize(avgYoungVoterPercent = mean(Ages.18.29.percent.estimate, na.rm=TRUE))
table02aByState$stateName <- tolower(table02aByState$stateName)
stateData <- map_data("state")
stateData <- left_join(stateData, table02aByState, by = c("region" = "stateName"))
ggplot(stateData, aes(x=long, y=lat))+ geom_polygon(aes(group=group, fill=avgYoungVoterPercent), color="white") + ggtitle("Average Percent Congressional District's Voters 18-29 by State") + scale_fill_distiller(name="Percent Composition", palette = "YlOrRd") + coord_map() + theme_void()
```

  The last graphs showed that the 18-29 age group is the most important age group to show up. So, this graph shows each State's average percent voters in that range. It however does not show a strict geographical pattern for these young voters. The Midwest does have a pattern of falling near the average, and those states with greater differences are coastal. Again, Florida can be noted as a unique state. A table earlier showed several of Florida's congressional districts having the greatest percentage in the nation, however this map shows that it has a very low average composition of voters 18-29. We can conclude that Florida has very diverse districts, some with high rates of young voters and others with high rates of older voters. 
  
# Geographic Analysis
# *Felix Baez*

```{r}
regions <- data.frame(State.name = c("Alabama", "Alaska", "Arizona", "Arkansas","California","Colorado","Connecticut","Delaware", "District of Columbia", "Florida", "Georgia","Hawaii","Idaho", "Illinois", "Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri" ,"Montana","Nebraska", "Nevada" ,"New Hampshire" , "New Jersey","New Mexico","New York", "North Carolina","North Dakota","Ohio", "Oklahoma","Oregon", "Pennsylvania","Rhode Island" ,"South Carolina" ,"South Dakota","Tennessee", "Texas" ,"Utah","Vermont","Virginia" , "Washington" , "West Virginia", "Wisconsin","Wyoming"), Region.name = c("Southeast", "Noncontiguous", "Southwest", "Southeast", "Pacific", "Rocky Mountains", "Northeast", "Southeast", "Southeast", "Southeast", "Southeast", "Noncontiguous", "Rocky Mountains", "Midwest", "Midwest", "Midwest", "Midwest", "Southeast", "Southeast", "Northeast", "Southeast", "Northeast", "Midwest", "Midwest", "Southeast", "Midwest", "Rocky Mountains", "Midwest", "Rocky Mountains", "Northeast", "Northeast", "Southwest", "Northeast", "Southeast", "Midwest", "Midwest", "Southwest", "Pacific", "Northeast", "Northeast", "Southeast", "Midwest", "Southeast", "Southwest", "Rocky Mountains", "Northeast", "Southeast", "Pacific", "Southeast", "Midwest", "Rocky Mountains"))

table01 <- read.csv("table01.csv")
votingByRegion <- table01 %>%
  inner_join(regions, by = "State.name") %>%
  group_by(Region.name) %>%
  summarise(votesCastForCongress = sum(Votes.cast.for.congressional.representative.for.the.November.6..2018.election, na.rm = TRUE), votingAgePopulation =sum(  Citizen.voting.age.population.estimate, na.rm = TRUE), votingRate = votesCastForCongress/votingAgePopulation)

mapping <- table01%>%
  inner_join(regions, by = "State.name") %>%
  inner_join(votingByRegion, by = "Region.name")

library(sf)
mapping <- mapping %>%
  mutate(Congressional.district=ifelse(Congressional.district=='At-large',0,Congressional.district))
map_data <- st_read('districts114.shp', quiet=TRUE) %>%
  mutate(
    STATENAME=ifelse(STATENAME=='District Of Columbia','District of Columbia',STATENAME),
    DISTRICT=ifelse(STATENAME=='District of Columbia',0,DISTRICT)
  ) %>%
  select(STATENAME, DISTRICT, geometry)
joined_data <- mapping %>%
  inner_join(map_data, by=c('State.name'='STATENAME','Congressional.district'='DISTRICT'))
joined_data %>%
  filter(!State.name %in% c('Hawaii','Alaska')) %>%
  ggplot(aes(fill=votingRate,geometry=geometry)) + geom_sf() + theme_void()
```

# Race Analysis
# *Ashley Sackpraseuth*

```{r}
race <- df5 %>%
  inner_join(df1, by = c("State.name", "Congressional.district", "Citizen.voting.age.population.estimate"))

race <- df5 %>% 
  select(State.abbreviation, State.name, Congressional.district, White.percent.total, Black.percent.total, Hispanic.percent.total,  Asian.percent.total, Two.or.more.percent.total)

race <- race %>% full_join(df1)

race[is.na(race)] <- 0 

race <- race %>%
  rename(White.voting.rate = White.percent.total, Black.voting.rate = Black.percent.total, Hispanic.voting.rate = Hispanic.percent.total, Asian.voting.rate = Asian.percent.total, Two.or.more.voting.rate = Two.or.more.percent.total)

race_summary <- race %>% pivot_longer(White.voting.rate:Two.or.more.voting.rate, names_to = "Race.ethnicity", values_to = "Percent") 

race_summary$Race.ethnicity <- factor(race_summary$Race.ethnicity, levels=c("White.voting.rate", "Black.voting.rate", "Hispanic.voting.rate", "Asian.voting.rate", "Two.or.more.voting.rate"))

# Overview
race_summary %>%
  group_by(Race.ethnicity) %>%
  summarise(mean = mean(Percent)) %>%
  mutate(mean = round(mean,1)) %>%
  arrange(desc(mean))

```

Before looking at the geographical distribution of voting partipcation by race, this table gives a brief introduction as to what to expect. White voters have the highest voting participation rate by far, making up, on average, 75.5% of a congressional district's total voting rate. This voting participation rate is then followed by Hispanic, Black, Asian, and Multiracial voters. 


# Do we want to move this map to the beginning of the report?
```{r}
map_base <- df1 %>%
  mutate(Congressional.district=as.numeric(ifelse(Congressional.district=='At-large',0,Congressional.district)))

map_data <- st_read('districts114.shp', quiet=TRUE) %>%
  mutate(
    STATENAME=ifelse(STATENAME=='District Of Columbia','District of Columbia',STATENAME),
    DISTRICT=as.numeric(ifelse(STATENAME=='District of Columbia',0,DISTRICT))
  ) %>%
  select(STATENAME, DISTRICT, geometry)


voting.rate_map <- map_base %>%
  filter(!State.name %in% c('Hawaii','Alaska')) %>%
  inner_join(map_data, by=c('State.name'='STATENAME','Congressional.district'='DISTRICT'))


voting.rate_map[is.na(voting.rate_map)] <- 0

df1 %>%
  summarise(mean = mean(Voting.rate.estimate))

total_voting_map <- voting.rate_map %>% 
  ggplot(aes(fill=Voting.rate.estimate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_gradient2(midpoint = 48.60069) + ggtitle("Total Voting Rate by Congressional District")

total_voting_map

```


```{r}
# Creating voting maps 
white_voting_map <- voting.rate_map %>% 
  ggplot(aes(fill=White.voting.rate, geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Blues', direction = 1) + ggtitle("White Voting Rate") + theme(legend.title=element_blank())

black_voting_map <- voting.rate_map %>% 
  ggplot(aes(fill=Black.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Purples', direction = 1)+ ggtitle("Black Voting Rate") + theme(legend.title=element_blank())

hispanic_voting_map <- voting.rate_map %>% 
  ggplot(aes(fill=Hispanic.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Greens', direction = 1) + ggtitle("Hispanic Voting Rate") + theme(legend.title=element_blank())

asian_voting_map <- voting.rate_map %>% 
  ggplot(aes(fill=Asian.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Reds', direction = 1) + ggtitle("Asian Voting Rate") + theme(legend.title=element_blank())

two.or.more_voting_map <- voting.rate_map %>% 
  ggplot(aes(fill=Two.or.more.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='YlGnBu', direction = 1 ) + ggtitle("Two or More Races Voting Rate") + theme(legend.title=element_blank())


#Displaying maps 
total_voting_map
grid.arrange(white_voting_map, black_voting_map, hispanic_voting_map, asian_voting_map, two.or.more_voting_map)
```

The Total Voting Rate by Congressional District map shows that the states in the northern half of the U.S. have above average voting turnout rates whereas the states in the southern half of the U.S. have below average voting turnout rates. The Total Voting Rate by Congressional District map looks very similar to the White Voting Rate map. The one area of the White Voting Rate map that has low voting rates (southeast corner of the U.S.), is the one area of the map where Black Voting Rates are very high. We are also seeing that Hispanic Voting Rates are highest in the southwest corner of the U.S. Asian Voting Rates are low across the map. Two or More Races Voting Rates are highest in Oklahoma and the west coast of the U.S. Overall, these maps tell us that Whites have a strong influence on overall voting rates.
The areas with the highest total voting rates had the highest White voting rates and the areas that had below average voting rates were the areas with the low White voting rates but high minority voting participation rates. 


```{r}
# Creating Census population maps 
states <- map_data("state")

census_race <- census_race %>%
  rename(REGION = 'ï..REGION')

states <- states %>%
  rename(state = region) %>%
  mutate(REGION = case_when(
    state == 'alabama' ~ 'South',
    state == 'arizona' ~ 'West',
    state == 'arkansas' ~ 'South',
    state == 'california' ~ 'West',
    state == 'colorado' ~ 'West',
    state == 'connecticut' ~ 'West',
    state == 'delaware' ~ 'South',
    state == 'florida' ~ 'South',
    state == 'georgia' ~ 'South',
    state == 'idaho' ~ 'West',
    state == 'illinois' ~ 'Midwest',
    state == 'indiana' ~ 'Midwest',
    state == 'iowa' ~ 'Midwest',
    state == 'kansas' ~ 'Midwest',
    state == 'kentucky' ~ 'South',
    state == 'louisiana' ~ 'South',
    state == 'maine' ~ 'Northeast',
    state == 'maryland' ~ 'South',
    state == 'massachusetts' ~ 'Northeast',
    state == 'michigan' ~ 'Midwest',
    state == 'minnesota' ~ 'Midwest',
    state == 'mississippi' ~ 'South',
    state == 'missouri' ~ 'South',
    state == 'montana' ~ 'West',
    state == 'nebraska' ~ 'Midwest',
    state == 'nevada' ~ 'West',
    state == 'new hampshire' ~ 'Northeast',
    state == 'new jersey' ~ 'Northeast',
    state == 'new mexico' ~ 'West',
    state == 'new york' ~ 'Northeast',
    state == 'north carolina' ~ 'South',
    state == 'north dakota' ~ 'Midwest',
    state == 'ohio' ~ 'Midwest',
    state == 'oregon' ~ 'West',
    state == 'oklahoma' ~ 'South',
    state == 'pennsylvania' ~ 'Northeast',
    state == 'rhode island' ~ 'Northeast',
    state == 'south carolina' ~ 'South',
    state == 'south dakota' ~ 'Midwest',
    state == 'tennessee' ~ 'South',
    state == 'texas' ~ 'South',
    state == 'utah' ~ 'West',
    state == 'vermont' ~ 'Northeast',
    state == 'virginia' ~ 'South',
    state == 'washington' ~ 'West',
    state == 'wisconsin' ~ 'Midwest',
    state == 'west virginia' ~ 'South',
    state == 'wyoming' ~ 'West'))

states <- states %>%
  left_join(census_race, by = "REGION")


white_pop_map <- states %>%
  ggplot(aes(x = long, y = lat, fill = White.percent)) + geom_polygon(aes(group=group)) + coord_map() +  geom_path(data = states, aes(x = long, y = lat, group = group), inherit.aes=FALSE, size=0.5, color='white') + theme_map() + scale_fill_distiller(palette='Blues', direction = 1) + ggtitle("White Pop. Percent") + theme(legend.title=element_blank(), legend.position = "right")

black_pop_map <- states %>%
  ggplot(aes(x = long, y = lat, fill = Black.percent)) + geom_polygon(aes(group=group)) + coord_map() +  geom_path(data = states, aes(x = long, y = lat, group = group), inherit.aes=FALSE, size=0.5, color='white') + theme_map() + scale_fill_distiller(palette='Purples', direction = 1)+ ggtitle("Black Pop. Percent") + theme(legend.title=element_blank(), legend.position = "right")

asian_pop_map <- states %>%
  ggplot(aes(x = long, y = lat, fill = Asian.percent)) + geom_polygon(aes(group=group)) + coord_map() +  geom_path(data = states, aes(x = long, y = lat, group = group), inherit.aes=FALSE, size=0.5, color='white') + theme_map() + scale_fill_distiller(palette='Reds', direction = 1) + ggtitle("Asian Pop. Percent") + theme(legend.title=element_blank(), legend.position = "right")

two.or.more_pop_map <- states %>%
  ggplot(aes(x = long, y = lat, fill = Two.or.more.percent)) + geom_polygon(aes(group=group)) + coord_map() +  geom_path(data = states, aes(x = long, y = lat, group = group), inherit.aes=FALSE, size=0.5, color='white') + theme_map()  + scale_fill_distiller(palette='YlGnBu', direction = 1 ) + ggtitle("Two or More Races Pop. Percent") + theme(legend.title=element_blank(), legend.position = "right")

grid.arrange(white_pop_map, black_pop_map, hispanic_pop_map, asian_pop_map, two.or.more_pop_map)

```

When comparing the geographical distribution of voters to the geographical distribution of citizens, few maps exhibit similar trends. Both the Black Population Percentage and Two or More Races Population Percentage map display the same trends seen in their voting rap maps, but does not hold true for the other three races analyzed. In both the Hispanic and Asian Population Percentage maps, we are seeing a high percentage of citizens in the west, but do not see their presence in their voting rate maps. This tells us that Hispanic and Asian citizens are not showing up to vote, especially on the west coast. The White Voting Rate makes it seem like there are more White citizens in the northern half of the U.S., but the Census data says otherwise. Because we are seeing a higher percentage of White presence in the Voting Rate map than the Population Percentage, we can conclude that White voters are the best at showing up to vote. 


# Poverty Analysis
# *Heather Moore*

```{r}
library(maps)
library(dplyr)
library(ggplot2)

Pov <- read.csv("table02b.csv")
Pov2 <- Pov %>% select(State.name,Congressional.district,Poverty.universe, Below.poverty.level.estimate, Below.poverty.level.percent.total, Citizen.voting.age.population.estimate)

names(Pov2)[names(Pov2) == "State.name"] <- "region"

Pov2$region <- tolower(Pov2$region)

Per_Coun <- Pov2 %>% select(region, Congressional.district, Below.poverty.level.percent.total) %>% filter(Below.poverty.level.percent.total > 20) %>% arrange(desc(Below.poverty.level.percent.total))
Per_Coun
```

The table above displays the congressional districts that have the highest percentage of voters under the poverty line. It appears the New York's 15th Congressional district has the most below poverty line voters at 30%. Following behind is Michigan's 13th and Kentucky's 5th districts at 24%. It's important to note that out of the top ten, four of the states are in the southern part of the United States. Also, both New York and Pennsylvania have 2 districts that fall under this classification.

```{r}
#Bar Chart 
BelPov <- Pov2 %>% group_by(region) %>% summarize(Pov_Uni = sum(Poverty.universe), Below_Pov = sum(Below.poverty.level.estimate)) 

BelPov$Percent_Below_Pov <- (BelPov$Below_Pov/BelPov$Pov_Uni) *100
BelPov2 <- BelPov %>% subset(Percent_Below_Pov > 14.6)

BelPov2 %>% ggplot(aes(region, Percent_Below_Pov)) + geom_bar(stat = 'identity', fill = 'blue') + labs(x = 'States', y = 'Percentage', title = 'Percentage of Voters Below the Poverty Line')  

```

After completing some analysis on the individual congressional districts, it will be meaningful to see if the same results would stand when looking at the states as a whole. The bar graph displays the states with the largest percentage of voters below the poverty line. In contrast from the table above, Mississippi has the the most below poverty line voters at almost 17% of voters participation falling into that category. An important pattern that is developing is that southern states tend to have the most voters that fall below the poverty line. 
```{r}
#map
library(RColorBrewer)

names(Pov2)[names(Pov2) == 'State.name'] <- 'region' 
State <- map_data('state')

map <- left_join(State, BelPov, by = 'region')
map %>% ggplot(aes(long, lat)) + geom_polygon(aes(group=group, fill = Percent_Below_Pov)) + scale_fill_distiller(name = "Percentage",palette = 'Blues') + theme_void() + labs(title = 'Percentage of Below the Poverty Line Voters by State') 
```

The final part of this analysis shows the correlation of below poverty line voters and the states. In this map,  the south holds the largest group of below poverty line voters. Which continues to proves the pattern that the south has most of the voters in that category.

## **Conclusion**
# Overall

    We looked at many factors and came to these specific conclusions. For education, the more educated a district is the higher voter participation there is, with bachelors degrees being the largest increase to voter participation. When considering age, there is a strong trend that congressional districts that have a higher percentage of voters 18-29, have a higher turnout in general. The inverse can be said for having a high percentage of voters 45-64. For Geography, ___. For Race, White voters have the highest turnout rate across the country, despite Census population data trends. For poverty, voter turnout is the highest in the Southern region of the United States. We believe that there are some forms of bias that influences these results ___.  So for the highest participation in a congressional district we are looking for a district with the highest percentage of bachelor degrees, and ___.
  
# Key take Aways
    1. Increasing education, specifically at the bachelor degree level is fastest way to increase voter participation from an education stand point.
    2.
    3.
    4.
    5. White voters have the highest voting participation rate. 
    6.Southern states have some of the highest turnout for voters who fall below the poverty line. 