---
title: "DS 202 Final Project"
author: "Kellan Bouwman, Adam Ford, Heather Moore, Brett White, Ashley Sackpraseuth, Felix Baez"
output: html_document
---

Introduction & Cleaning
*Brett White*
    
     
```{r}
# tidyverse
library(tidyverse)
# readxl
library(readxl)
df1 <- read.csv('table01.csv')
df2 <- read.csv('table02a.csv')
df3 <- read.csv('table02b.csv')
df4 <- read.csv('table02c.csv')
df5 <- read.csv('table02d.csv')
# head(df1)
# head(df2)
# head(df3)
# head(df4)
# head(df5)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(lubridate)
```

## Import and Rename Columns
```{r}
table01 <- read.csv("table01.csv")
table01 <- table01 %>%
  rename(
    stateAbb = State.abbreviation,
    stateName = State.name,
    congressDist = Congressional.district,
    votesCastForCongress = Votes.cast.for.congressional.representative.for.the.November.6..2018.election,
    votingAgePopulation = Citizen.voting.age.population.estimate,
    votingAgePopulationMOE = Citizen.voting.age.population.Margin.of.Error,
    votingRate = Voting.rate.estimate,
    votingRateMOE = Voting.rate.Margin.of.Error
  )
head(table01, 5)
table02a <- read.csv("table02a.csv")
table02a <- table02a %>%
  rename(
    stateAbb = State.abbreviation,
    stateName = State.name,
    congressDist = Congressional.district,
    votingAgePopulation = Citizen.voting.age.population.estimate,
    votingAgePopulationMOE = Citizen.voting.age.population.MOE
  )
table02b <- read.csv("table02b.csv")
table02b <- table02b %>%
  rename(
    stateAbb = State.abbreviation,
    stateName = State.name,
    congressDist = Congressional.district,
    votingAgePopulation = Citizen.voting.age.population.estimate,
    votingAgePopulationMOE = Citizen.voting.age.population.MOE
  )
```


Education Analysis
*Kellan Bouwman*
  When looking at the specific educational categories there does not seem to be any specific trend. If we are lumping them into a group though, the more educated they are the higher voting percentage for a congressional district. This is further seen when looking at Highschool and bachelor degree estimates compared with the national average, the more educate a district is the more participation there is. With Highschool graduates the estimate range has a lot more variability, but has a similar estimating ability as bachelor degrees. 

```{r}
# head(df4)
eduPol <- df4 %>% select(State.abbreviation, State.name, Congressional.district, Citizen.voting.age.population.estimate, Less.grade.9.estimate, Grade.9.12.no.diploma.estimate, High.school.graduate.estimate, Some.college.no.degree.estimate, Associate.degree.estimate, Bachelor.degree.estimate, Graduate.degree.estimate, High.school.or.more.estimate, Bachelor.degree.or.more.estimate)

eduPol %>% full_join(df1) -> eduPol

eduPol %>% pivot_longer(Less.grade.9.estimate:Graduate.degree.estimate, names_to = "Education", values_to = "Amount") -> eduPol
#head(eduPol)

eduPol %>%
  mutate(PercentageEdu = (Amount / Citizen.voting.age.population.estimate) * 100 ) -> eduPol

eduPol %>%
  group_by(Education) %>%
  summarise(mean = mean(PercentageEdu, na.rm=TRUE))

highSchoolPlus <- mean((eduPol$High.school.or.more.estimate / eduPol$Citizen.voting.age.population.estimate) * 100)
BachPlus <- mean((eduPol$Bachelor.degree.or.more.estimate / eduPol$Citizen.voting.age.population.estimate) * 100)

eduPol %>%
  mutate(nationalAvgDiffHighGrad = ((High.school.or.more.estimate / Citizen.voting.age.population.estimate) * 100) - highSchoolPlus, nationalAvgDiffBach = ((Bachelor.degree.or.more.estimate / Citizen.voting.age.population.estimate) * 100) - BachPlus, highschoolgradper = ((High.school.or.more.estimate / Citizen.voting.age.population.estimate) * 100), bachgradper = ((Bachelor.degree.or.more.estimate / Citizen.voting.age.population.estimate) * 100)) -> eduPol

#Overall Comparisons
ggplot(eduPol, aes(x=PercentageEdu, y=Voting.rate.estimate)) + geom_point() + geom_smooth(method='lm') + facet_wrap(~Education, scales = "free") + ggtitle("Voter Rate based off Highest level of education") + xlab("Percentage of Population") + ylab("Voter Participation Percentage")

#Highschool Graduation corralation with voter rate
ggplot(eduPol, aes(x=nationalAvgDiffHighGrad, y=Voting.rate.estimate)) + geom_point() + geom_smooth(method='lm') + ggtitle("Voter Rate & difference Highschool or more education & the National Average") + xlab("Percentage of Population, difference with the National Average") + ylab("Voter Participation Percentage")

#Bachelors Degrees holders
ggplot(eduPol, aes(x=nationalAvgDiffBach, y=Voting.rate.estimate)) + geom_point() + geom_smooth(method='lm') + ggtitle("Voter Rate & difference Bachelors Degrees or more education & National Average") + xlab("Percentage of Population, difference with the National Average") + ylab("Voter Participation Percentage")

#Map

#head(eduPol)
```

Age Analysis
*Adam Ford*

```{r}
table02a <- table01 %>%
  select(stateAbb, congressDist, votingRate) %>%
  inner_join(table02a, by = c('stateAbb', 'congressDist'))
mean18.29 <- mean(table02a$Ages.18.29.percent.estimate)
mean30.44 <- mean(table02a$Ages.30.44.percent.estimate)
mean45.64 <- mean(table02a$Ages.45.64.percent.estimate)
mean65.up <- mean(table02a$Ages.65.up.percent.estimate)
# mean18.29
# mean30.44
# mean45.64
# mean65.up
table02a$dif18.29 <- mean18.29 - table02a$Ages.18.29.percent.estimate
table02a$dif30.44 <- mean30.44 - table02a$Ages.30.44.percent.estimate
table02a$dif45.64 <- mean45.64 - table02a$Ages.45.64.percent.estimate
table02a$dif65.up <- mean65.up - table02a$Ages.65.up.percent.estimate
```

### 18 - 29
```{r}
table02a %>% 
  select(stateName, congressDist, dif18.29, votingRate) %>%
  arrange(desc(dif18.29)) %>%
  head(5)
```

### 30 - 44
```{r}
table02a %>%
  select(stateName, congressDist, dif30.44, votingRate) %>%
  arrange(desc(dif30.44)) %>%
  head(5)
```

### 45 - 64
```{r}
table02a %>%
  select(stateName, congressDist, dif45.64, votingRate) %>%
  arrange(desc(dif45.64)) %>%
  head(5)
```

### 65 and up
```{r}
table02a %>%
  select(stateName, congressDist, dif65.up, votingRate) %>%
  arrange(desc(dif65.up)) %>%
  head(5)
```

  These tables show the congressional districts that have the nationwide greatest percentage of voters in the stated age groups. A notable state on these lists is Florida, it dominates the maximums for young voters. A pattern is beginning to show in the Voting Rates as well. We see that the states with high populations of young voters also have a High Voting Rate.
```{r}
ggplot(table02a, aes(x=dif18.29, y=votingRate)) +geom_point() + geom_smooth(method='lm') + xlab("Difference in Percent Composition of Voters Ages 18-29 compared to the National Average") + ylab("Voting Rate")+ ggtitle("Difference in the 18-29 age group versus Voting Rate in Congressional Districts")

ggplot(table02a, aes(x=dif45.64, y=votingRate)) +geom_point() + geom_smooth(method='lm')  + xlab("Difference in Percent Composition of Voters Ages 45-64 compared to the National Average") + ylab("Voting Rate")+ ggtitle("Difference in the 45-64 age group versus Voting Rate in Congressional Districts")
```

  These two graphs show two starkly different relationships between Difference of Composition of Voters Ages 18-29 and 45-64 to the National Averages versus the respective Voting Rates. In more laymans terms, congressional districts thats has a higher percentage of voters 18 - 29 than the nation, tend to have better voter turnouts. The opposite can be said about the age group of 45 - 64. Therefore, we can conclude that the rate of voters 18-29 is an important factor of what a congressional district's voting rate will be.
  
```{r}
library("maps")
table02aByState <- table02a %>%
  group_by(stateName) %>%
  summarize(avgYoungVoterPercent = mean(Ages.18.29.percent.estimate, na.rm=TRUE))
table02aByState$stateName <- tolower(table02aByState$stateName)
stateData <- map_data("state")
stateData <- left_join(stateData, table02aByState, by = c("region" = "stateName"))
ggplot(stateData, aes(x=long, y=lat))+ geom_polygon(aes(group=group, fill=avgYoungVoterPercent), color="white") + ggtitle("Average Percent Congressional District's Voters 18-29 by State") + scale_fill_distiller(name="Percent Composition", palette = "YlOrRd") + coord_map() + theme_void()
```

  This graph does not show a strict geographical pattern for young voters. The midwest does all fall near the average, and those states with greater difference's are coastal. Again, Florida can be noted as a unique state. A table earlier showed several of Florida's congressional districts having the greatest percentage in the nation, however this map shows that it has a very low average composition of voters 18-29. We can conclude that Florida has very diverse districts, some immensely heavy in young voters, and others heavily older. 
  
Geographic Analysis
*Felix Baez*

```{r}
regions <- data.frame(State.name = c("Alabama", "Alaska", "Arizona", "Arkansas","California","Colorado","Connecticut","Delaware", "District of Columbia", "Florida", "Georgia","Hawaii","Idaho", "Illinois", "Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri" ,"Montana","Nebraska", "Nevada" ,"New Hampshire" , "New Jersey","New Mexico","New York", "North Carolina","North Dakota","Ohio", "Oklahoma","Oregon", "Pennsylvania","Rhode Island" ,"South Carolina" ,"South Dakota","Tennessee", "Texas" ,"Utah","Vermont","Virginia" , "Washington" , "West Virginia", "Wisconsin","Wyoming"), Region.name = c("Southeast", "Noncontiguous", "Southwest", "Southeast", "Pacific", "Rocky Mountains", "Northeast", "Southeast", "Southeast", "Southeast", "Southeast", "Noncontiguous", "Rocky Mountains", "Midwest", "Midwest", "Midwest", "Midwest", "Southeast", "Southeast", "Northeast", "Southeast", "Northeast", "Midwest", "Midwest", "Southeast", "Midwest", "Rocky Mountains", "Midwest", "Rocky Mountains", "Northeast", "Northeast", "Southwest", "Northeast", "Southeast", "Midwest", "Midwest", "Southwest", "Pacific", "Northeast", "Northeast", "Southeast", "Midwest", "Southeast", "Southwest", "Rocky Mountains", "Northeast", "Southeast", "Pacific", "Southeast", "Midwest", "Rocky Mountains"))

table01 <- read.csv("table01.csv")
votingByRegion <- table01 %>%
  inner_join(regions, by = "State.name") %>%
  group_by(Region.name) %>%
  summarise(votesCastForCongress = sum(Votes.cast.for.congressional.representative.for.the.November.6..2018.election, na.rm = TRUE), votingAgePopulation =sum(  Citizen.voting.age.population.estimate, na.rm = TRUE), votingRate = votesCastForCongress/votingAgePopulation)

mapping <- table01%>%
  inner_join(regions, by = "State.name") %>%
  inner_join(votingByRegion, by = "Region.name")

library(sf)
mapping <- mapping %>%
  mutate(Congressional.district=ifelse(Congressional.district=='At-large',0,Congressional.district))
map_data <- st_read('districts114.shp', quiet=TRUE) %>%
  mutate(
    STATENAME=ifelse(STATENAME=='District Of Columbia','District of Columbia',STATENAME),
    DISTRICT=ifelse(STATENAME=='District of Columbia',0,DISTRICT)
  ) %>%
  select(STATENAME, DISTRICT, geometry)
joined_data <- mapping %>%
  inner_join(map_data, by=c('State.name'='STATENAME','Congressional.district'='DISTRICT'))
joined_data %>%
  filter(!State.name %in% c('Hawaii','Alaska')) %>%
  ggplot(aes(fill=votingRate,geometry=geometry)) + geom_sf() + theme_void()
```

Race Analysis
*Ashley Sackpraseuth*

```{r}
race <- df5 %>%
  inner_join(df1, by = c("State.name", "Congressional.district", "Citizen.voting.age.population.estimate"))

race <- df5 %>% 
  select(State.abbreviation, State.name, Congressional.district, White.percent.total, Black.percent.total, Hispanic.percent.total,  Asian.percent.total, Two.or.more.percent.total)

race %>% full_join(df1) -> race

race[is.na(race)] <- 0 

race <- race %>%
  rename(White.voting.rate = White.percent.total, Black.voting.rate = Black.percent.total, Hispanic.voting.rate = Hispanic.percent.total, Asian.voting.rate = Asian.percent.total, Two.or.more.voting.rate = Two.or.more.percent.total)

race_scatterplot <- race %>% pivot_longer(White.voting.rate:Two.or.more.voting.rate, names_to = "Race.ethnicity", values_to = "Percent") 

race_scatterplot$Race.ethnicity <- factor(race_scatterplot$Race.ethnicity, levels=c("White.voting.rate", "Black.voting.rate", "Hispanic.voting.rate", "Asian.voting.rate", "Two.or.more.voting.rate"))

race_scatterplot %>%
  group_by(Race.ethnicity) %>%
  summarise(mean = mean(Percent))


ggplot(race_scatterplot, aes(x=Voting.rate.estimate, y=Percent)) + geom_point() + geom_smooth(method='lm') + facet_wrap(~Race.ethnicity, scales = "free_y") + xlim(27, 72) + ggtitle("Race/Ethnicity Voting Rates vs. Total Congressional District Voting Rate") + xlab("Total Congressional District Voting Rate") + ylab("Race/Ethnicity Voting Rate")
```


Race Voting Maps
*Ashley Sackpraseuth*

```{r}
library(sf)

map_base <- race %>%
  mutate(Congressional.district=as.numeric(ifelse(Congressional.district=='At-large',0,Congressional.district)))

map_data <- st_read('districts114.shp', quiet=TRUE) %>%
  mutate(
    STATENAME=ifelse(STATENAME=='District Of Columbia','District of Columbia',STATENAME),
    DISTRICT=as.numeric(ifelse(STATENAME=='District of Columbia',0,DISTRICT))
  ) %>%
  select(STATENAME, DISTRICT, geometry)


voting.rate_map <- map_base %>%
  filter(!State.name %in% c('Hawaii','Alaska')) %>%
  inner_join(map_data, by=c('State.name'='STATENAME','Congressional.district'='DISTRICT'))


voting.rate_map[is.na(voting.rate_map)] <- 0

race %>%
  summarise(mean = mean(Voting.rate.estimate))

voting.rate_map %>% 
  ggplot(aes(fill=Voting.rate.estimate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_gradient2(midpoint = 48.60069) + ggtitle("Total Voting Rate by Congressional District")


# Map of Voting Rate Race Breakdown

#install.packages(gridExtra)
library(gridExtra)

white_map <- voting.rate_map %>% 
  ggplot(aes(fill=White.voting.rate, geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Blues', direction = 1) + ggtitle("White Voting Rate") + theme(legend.title=element_blank())

black_map <- voting.rate_map %>% 
  ggplot(aes(fill=Black.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Purples', direction = 1)+ ggtitle("Black Voting Rate") + theme(legend.title=element_blank())

hispanic_map <- voting.rate_map %>% 
  ggplot(aes(fill=Hispanic.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Greens', direction = 1) + ggtitle("Hispanic Voting Rate") + theme(legend.title=element_blank())

asian_map <- voting.rate_map %>% 
  ggplot(aes(fill=Asian.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='Reds', direction = 1) + ggtitle("Asian Voting Rate") + theme(legend.title=element_blank())

two.or.more_map <- voting.rate_map %>% 
  ggplot(aes(fill=Two.or.more.voting.rate,geometry=geometry)) + geom_sf() + theme_void() + scale_fill_distiller(palette='YlGnBu', direction = 1 ) + ggtitle("Two or More Races Voting Rate") + theme(legend.title=element_blank())

grid.arrange(white_map, black_map, hispanic_map, asian_map, two.or.more_map)
```

Both the scatterplot and White Voting Rate map show that Whites have the highest voting participation rate out of the five racial/ethnic groups analysed. They're followed by Hispanics, Blacks, Asians, and then those who are Two or More races. The Total Voting Rate by Congressional District map shows that the states in the northern half of the U.S. have above average voting turnout rates whereas the states in the southern half of the U.S. have below average voting turnout rates. The Total Voting Rate by Congressional District map looks very similar to the White Voting Rate map. The one area of the White Voting Rate map that has low voting rates (southeast corner of the U.S.), is the one area of the map where Black Voting Rates are very high. We are also seeing that Hispanic Voting Rates are highest in the southwest corner of the U.S. Asian Voting Rates are low across the map. Two or More Races Voting Rates are highest in Oklahoma and the west coast of the U.S. Overall, these maps tell us that Whites have a strong influence on overall voting rates.
The areas with the highest total voting rates had the highest White voting rates and the areas that had below average voting rates were the areas with the low White voting rates but high minority voting participation rates. 



Poverty Analysis
*Heather Moore*

```{r}
library(maps)
library(dplyr)
library(ggplot2)

Pov <- read.csv("table02b.csv")
Pov2 <- Pov %>% select(State.name,Congressional.district,Poverty.universe, Below.poverty.level.estimate, Below.poverty.level.percent.total, Citizen.voting.age.population.estimate)

names(Pov2)[names(Pov2) == "State.name"] <- "region"

Pov2$region <- tolower(Pov2$region)

Per_Coun <- Pov2 %>% select(region, Congressional.district, Below.poverty.level.percent.total) %>% filter(Below.poverty.level.percent.total > 20) %>% arrange(desc(Below.poverty.level.percent.total))
Per_Coun
```

The table above displays the congressional districts that have the highest percentage of voters under the poverty line. It appears the New York's 15th Congressional district has the most below poverty line voters at 30%. Following behind is Michigan's 13th and Kentucky's 5th districts at 24%. It's important to note that out of the top ten, four of the states are in the southern part of the United States. Also, both New York and Pennsylvania have 2 districts that fall under this classification.

```{r}
#Bar Chart 
BelPov <- Pov2 %>% group_by(region) %>% summarize(Pov_Uni = sum(Poverty.universe), Below_Pov = sum(Below.poverty.level.estimate)) 

BelPov$Percent_Below_Pov <- (BelPov$Below_Pov/BelPov$Pov_Uni) *100
BelPov2 <- BelPov %>% subset(Percent_Below_Pov > 14.6)

BelPov2 %>% ggplot(aes(region, Percent_Below_Pov)) + geom_bar(stat = 'identity', fill = 'blue') + labs(x = 'States', y = 'Percentage', title = 'Percentage of Voters Below the Poverty Line')  

```

After completing some analysis on the individual congressional districts, it will be meaningful to see if the same results would stand when looking at the states as a whole. The bar graph displays the states with the largest percentage of voters below the poverty line. In contrast from the table above, Mississippi has the the most below poverty line voters at almost 17% of voters participation falling into that category. An important pattern that is developing is that southern states tend to have the most voters that fall below the poverty line. 
```{r}
#map
library(RColorBrewer)

names(Pov2)[names(Pov2) == 'State.name'] <- 'region' 
State <- map_data('state')

map <- left_join(State, BelPov, by = 'region')
map %>% ggplot(aes(long, lat)) + geom_polygon(aes(group=group, fill = Percent_Below_Pov)) + scale_fill_distiller(palette = 'Blues') + theme_void() + labs(title = 'Percentage of Below the Poverty Line Voters by State') + coord_map()
```

The final part of this analysis shows the correlation of below poverty line voters and the states. In this map,  the south holds the largest group of below poverty line voters. Which continues to proves the pattern that the south has most of the voters in that category.

Conclusion
  
  We look at many factors and came to these specific conclusions. For education, the more educated a district is the higher voter participation there is, with bachelors degrees being the largest increase to voter participation. For age, ___. For Geography, ___. For Race, ___. For poverty, ___. We believe that there are some forms of bias that influences these results ___.  So for the highest participation in a congressional district we are looking for a district with the highest percentage of bachelor degrees, and ___.